---
title: "Chapter 1 - Quickstart Guide"
author: "Gabriel Odom"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
  word_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Quickstart Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE,
                      cache = FALSE,
                      comment = "#>")
```


# Overview
This guide will serve as a brief introduction to pathway significance testing with the `pathwayPCA` package. We will discuss four steps. For detailed discussion of these four steps, see the appropriate vignettes. This vignette is the introduction and overview chapter in the "Pathway Significance Testing with `pathwayPCA`" workflow. 

0. Download Packages
1. Import Data ([*vignette*](https://gabrielodom.github.io/pathwayPCA/articles/C2-Importing_Data.html))
2. Create `Omics` Data Object ([*vignette*](https://gabrielodom.github.io/pathwayPCA/articles/C3-Create_Omics_Objects.html))
3. Test Pathway Significance ([*vignette*](https://gabrielodom.github.io/pathwayPCA/articles/C4-Methods_Walkthrough.html))
4. Inspect Results

Before we get started, you need the `pathwayPCA` package to run your analysis. You can install it from GitHub. If you want your analysis to be performed with parallel computing, you will need a package to help you. We recommend the `parallel` package. We also recommend the `tidyverse` package to help you run some of the examples in these vignettes. 
```
devtools::install_github("gabrielodom/pathwayPCA")
```
```{r}
library(pathwayPCA)
library(tidyverse)
library(parallel)
```

*******************************************************************************

</br>


# Import Data
This section is a quick overview of the material covered in the [Import and Tidy Data](https://gabrielodom.github.io/pathwayPCA/articles/C2-Importing_Data.html) vignette. We will cover three data import steps.

### Import `.gmt` Files
Use the `read_gmt` function. Create your own with the `create_pathwaySet` function.

### Import and Tidy Assay Data
Use the `read_csv` function from the `readr` package. Tidy the assay with the `transpose_assay` function.

### Import Patient Info
Use the `read_csv` function to import the patient data, then the `inner_join` function from the `dplyr` package to match the assay measurements to patient information by subject identifier.

*******************************************************************************

</br>


# Create an `Omics` Data Object
This section is a quick overview of the material covered in the [Create an `Omics*` Data Container](https://gabrielodom.github.io/pathwayPCA/articles/C3-Create_Omics_Objects.html) vignette.

### Create an Object
Create a data container specific to survival, regression, or categorical responses.

### Inspect the Object Contents
Print the object to the screen to see a summary of the data contained therein.

### Modify Object Contents
Use accessor functions to extract, edit, or replace data contained in the object.

*******************************************************************************

</br>


# Test Pathways for Significance
This section is a quick overview of the material covered in the "AES-PCA" and "Supervised PCA" sections of the [Test Pathway Significance](https://gabrielodom.github.io/pathwayPCA/articles/C4-Methods_Walkthrough.html) vignette.


### AES-PCA
Perform AES-PCA on the object with the `AESPCA_pVals` function.

### Supervised PCA
Perform Supervised PCA on the object with the `superPCA_pVals` function.

*******************************************************************************

</br>


# Inspect Results
This section is a quick overview of the material covered in the "Analyze the Results" section of the [Test Pathway Significance](https://gabrielodom.github.io/pathwayPCA/articles/C4-Methods_Walkthrough.html) vignette.

### Graph of Top Pathways
Use `ggplot2` to create summary graphics of the analysis results.

### Extract Genes from the Top Pathways
Use the `topGenes` function to "score" the genes contained in the top significant pathways.
