---
title: "Supervised PCA Walkthrough with pathwayPCA"
author: "Gabriel Odom"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This will provide the steps to perform pathway-level supervised PCA analysis, from data entry to pathway $p$-value graphics. This vignette will give three examples: survival, regression, and binary responses.

## Workflow Setup and Data
Before we begin, make sure you have the latest version of the package. In order to install a package from GitHub, you will need the `devtools::` package (https://github.com/r-lib/devtools) and either [Rtools](https://cran.r-project.org/bin/windows/Rtools/) (for Windows) or [Xcode](https://developer.apple.com/xcode/) (for Mac). Then you can install the development version of the [`pathwayPCA` package](https://github.com/gabrielodom/pathwayPCA) from [GitHub](https://github.com/):
```{r install_from_github, cache = TRUE}
devtools::install_github("gabrielodom/pathwayPCA")
library(pathwayPCA)
```
